---
title: "HW2 Telemarketing"
author: "Sameer Gurijala"
date: "11/08/2020"
output:
  html_document:
    toc: true
    theme: readable
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Downloading and Prepping the Data

```{r}
#Downloading and Prepping the Data
tele <- read.csv("tele.csv", stringsAsFactors = TRUE)
summary(tele)

#We are deleting the "duration" variable because it is an after the fact measurement. We only should be using variables that we know before the call
tele$duration <- NULL

# Deleting the column X
tele$X <- NULL

# Changing pdays to a dummy and deleting pdays
tele$pdaysdummy <- ifelse(tele$pdays == 999, 0, 1)
tele$pdays <- NULL

str(tele)
```

## Getting Data Ready for Analysis

```{r}
# Using model.matrix to convert all the factors to dummy variables
# We are converting all of the factors into dummy variables as the input into ANN has to be numeric

telemm <- as.data.frame(model.matrix(~.-1,tele))
str(telemm)

# Randomize the rows in the data (shuffling the rows)
set.seed(12345)
tele_random <- telemm[sample(nrow(telemm)),]

#Normalize the data
normalize <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)))
}

# we are going to normalize everything 
tele_norm <- as.data.frame(lapply(tele_random, normalize))
```


## Getting Train and Test Samples

```{r}
# Selects 10000 random rows for test data
set.seed(12345)
test_set <- sample(1:nrow(tele_norm), 10000) 
# Depending on R-version and computer, different rows may be selected. 
# If that happens, results are different. 

# Create a train set and test set
#First the predictors - all columns except the yyes column
tele_train <- tele_norm[-test_set, ]
tele_test <- tele_norm[test_set, ]
```

> Now you are ready to build your ANN model. Feel free to modify the data load, cleaning and preparation code above as per your preference.

## Logistic Regression Model

```{r}

tele_model_LR <- glm(yyes ~ ., data = tele_train, family = binomial(link = "logit"))
summary(tele_model_LR)

```

### Stepwise Regression to find significant variables

```{r}
tele_stepmodel <- step(tele_model_LR, direction = "backward")
summary(tele_stepmodel)
```

### Prediction based on Step Model

```{r}
pred_step <- predict(tele_stepmodel, tele_test, type = "response")
summary(pred_step)
pred_cat <- ifelse(pred_step >= 0.5, 1, 0)
table(pred_cat)
300/9700
table(tele_train$yyes)
3540/27648
```

### Evaluation of Model

```{r}
library(caret)
confusionMatrix(as.factor(pred_cat), as.factor(tele_test$yyes), positive = "1")
```


## ANN Model

```{r}
library(neuralnet)

# simple ANN with only a single hidden neuron
tele_model_ANN <- neuralnet(formula = yyes ~ .,
                              data = tele_train)


# visualize the network topology
plot(tele_model_ANN)

## Step 4: Evaluating model performance ----
# obtain model results
ANNmodel_results <- compute(tele_model_ANN, tele_test[,-29])
# obtain predicted strength values
ANNsubscribed <- ANNmodel_results$net.result
# examine the correlation between predicted and actual values
cor(ANNsubscribed, tele_test$yyes)

predicted_neural01_cat <- ifelse(ANNsubscribed > 0.5, 1 , 0)
confusionMatrix(as.factor(predicted_neural01_cat), as.factor(tele_test$yyes), positive = "1")

```

## ANN with Hidden 

```{r}

tele_model_ANN_Hidden4 <- neuralnet(formula = yyes ~.
                              ,data = tele_train, hidden = 2)
plot(tele_model_ANN_Hidden4)

## Step 4: Evaluating model performance ----
# obtain model results
ANNmodel_resultsHidden4 <- compute(tele_model_ANN_Hidden4, tele_test[,-29])
# obtain predicted strength values
ANNsubscribedHidden4 <- ANNmodel_resultsHidden4$net.result
# examine the correlation between predicted and actual values
cor(ANNsubscribedHidden4, tele_test$yyes)

predicted_neural01_catHidden4 <- ifelse(ANNsubscribedHidden4 > 0.5, 1 , 0)
confusionMatrix(as.factor(predicted_neural01_catHidden4), as.factor(tele_test$yyes), positive = "1")
```
## Conclusion
Among the three models, the ANN with 2 hidden layers gives the highest accuracy with 0.9244

## Collaborated with Jatin Sindhu